## HW1

## Case study 1:

Objective: estimate audience size for the Wharton show

Approach: estimate the proportion of the Wharton listeners to that of the Sirius listeners, p, so that we will come up with an audience size estimate of approximately 51.6 million times p

```{r}
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(tidyverse, skimr, nycflights13, gapminder, ggthemes, ggpubr, data.table, plotly)
```

Data prep

```{r, eval = F}
results <- read.csv("data/Survey_results_final.csv")
```

```{r}
cleaned_results <-
results %>% 
  select(Answer.Age, Answer.Gender, Answer.Education, Answer.HouseHoldIncome, Answer.Sirius.Radio, Answer.Wharton.Radio, WorkTimeInSeconds) %>% 
  rename(age = Answer.Age, gender = Answer.Gender, education = Answer.Education, income = Answer.HouseHoldIncome, sirius = Answer.Sirius.Radio, wharton = Answer.Wharton.Radio, worktime = WorkTimeInSeconds) %>% 
  na.omit() #%>% 
  #filter(age != "", gender != "", education != "", income != "", sirius != "", wharton != "", worktime != "")
```

```{r}
cleaned_results %>%
  filter(worktime <= 5)
```

A work time of less than or equal to 5 seconds means that people did not spend enough time to look through the survey at all, so those responses can be disregarded.

```{r}
cleaned_results %>%
  filter(age == "")
```

```{r}
# wharton listeners
cleaned_results %>%
  group_by(wharton) %>%
  summarise(wharton_listeners = n())
```

```{r}
# sirius listeners 
cleaned_results %>% 
  group_by(sirius) %>% 
  summarise(sirius_listeners = n())
```
