---
title: "STAT571_hw1"
author: "Ruolan Li"
date: "2023-01-20"
output: html_document
---

# Case study 3: Major League Baseball

**Data input**

```{r data input}
datapay_wide <- read.csv("/Users/ruolanli/Downloads/MLPayData_Total.csv")
datapay_long <- read.csv("/Users/ruolanli/Downloads/baseball.csv")
```


## 4.1 EDA: Relationship between payroll changes and performance

### (a)
```{r compare difference and log difference}
library(dplyr)
library(ggplot2)
#plot variable payroll directly
p1 <- datapay_long %>%
  ggplot(aes(x = payroll)) +
  geom_density()

#plot log transformed payroll
p2 <- datapay_long %>%
  ggplot(aes(x = log(payroll))) +
  geom_density()

ggarrange(p1,p2)

```

\ As the density plot shown, the variable payroll is right skewed. After the log transformation, the density plot is much less skewed. Therefore, it is more appropriate to use log difference.


### (b)
```{r create diff_log variable}
#log transform difference
datapay_long = datapay_long %>%
  group_by(team) %>%
  mutate(diff_log = log(payroll) - log(lag(payroll)))
head(datapay_long)
```


### (c)
```{r create long table}
datapay_long = datapay_long %>%
  select(team, year, diff_log, win_pct)
head(datapay_long)
```

## 4.2 Exploratory questions
### (a)
```{r highest increase in payroll between years 2010 and 2014}
payroll <- datapay_long %>%
  group_by(team) %>%
  filter(year %in% c(2010:2014)) %>%
  summarise(diff_sum = sum(diff_log)) %>% #log(2014)-log(2010) = sum of log(2014)-log(2013)+log(2013)-log(2012)...
  arrange(desc(diff_sum))
head(payroll)
```

\ Los Angeles Dodgers, Washington Nationals, San Diego Padres, Texas Rangers and San Francisco Giants had ighest increase in their payroll between years 2010 and 2014.


### (b)
```{r biggest percentage gain in wins}
win <- datapay_wide %>%
  rename("team" = names(.)[1]) %>%
  select(team, paste0("X", c(2014:2010))) %>%
  group_by(team) %>%
  mutate(pct = sum(c_across(X2014:X2010)) / X2010) %>% #calculate percentage between 2010 and 2014
  arrange(desc(pct))
head(win)
```

\ Pittsburgh Pirates had the biggest percentage gain in wins between 2010 and 2014.


## 4.3 Do log increases in payroll imply better performance?
```{r test higher increases in payroll lead to increased performance}
#scatterplot
datapay_agg <- payroll %>%
  inner_join(win, by = c("team" = "team")) %>%
  select(team,diff_sum,pct)
datapay_agg


datapay_agg %>%
  ggplot(aes(x = diff_sum, y = pct)) +
  geom_point(color = "blue", size = 3) +
  geom_text(aes(label = team), size = 3) +
  labs(title = "Win Percentage vs. Log Increase in Payroll",
       x = "Log_increase_in_payroll",
       y = "Win_pct")


```
```{r}
#Least Squared Lines
datapay_agg %>%
  ggplot(aes(x = diff_sum, y = pct))+
  geom_point(size = 3)+
  geom_smooth(method = "lm", formula = y~x, color = "blue")+
  labs(title = "Win Percentage vs. Log Increase in Payroll",
       x = "Log_increase_in_payroll",
       y = "Win_pct")+
  theme_bw()
```



```{r}
#linear relationship
fit = lm(pct~diff_sum, data = datapay_agg)
summary(fit)

#correaltion
cor(datapay_agg$diff_sum, datapay_agg$pct)
```

\ No, there is no strong evidence to support the hypothesis that higher increases in payroll on the log scale lead to increased performance. 

\ In the scatter plot, the data points do not cluster tightly. 

\ R-squared in the linear model shows the linear regression model did not fit the data well, indicating no strong linear relationship between log increases in payroll and win percentage.

\ The correlation coefficient is 0.294, indicating a low correlation.


## 4.4 Comparison



